Pattern matching for\\ $(213,231)$-avoiding permutations

### 2015-04-19 article.tex
typo

\begin{abstract}
	Given permutations $\sigma \in S_k$ and $\pi \in S_n$, the
  \emph{pattern containment} problem is to decide whether $\pi$ contains
  $\sigma$ as an order-isomorphic subsequence.
	We give a linear-time algorithm in case both $\pi$ and $\sigma$ avoid
	$213$ and $231$.
	For the special where only $\sigma$ avoids $213$ and $231$, we present a
	$O(kn^2)$ time algorithm. We extend our research to $(213,231)$-avoiding
	bivincular pattern and present a $O(kn^3)$ time algorithm.
\end{abstract}

@@

This section is devoted to presenting a fast algorithm for deciding
$\sigma \leq \pi$
in case both $\pi$ and $\sigma$ are $(213,231)$-avoiding.
We begin with an easy but crucial structure lemma.

@@

\begin{proof}
Any other initial element would serve as a `$2$' in either a
$231$ or $213$ with $1$ and $n$ as the `$1$' and `$3$' respectively.
\end{proof}

Notice that Lemma~\ref{lemma:first element is 1 or n} gives as an
immediate consequence that the number of $(213,231)$-avoiding $n$-permutations
is $2^{n-1}$ for $n\geq 1$ \cite{Simion:Schmidt:EJC:1985}.
The above lemma also gains in interest in the form of the following corollary.

\begin{corollary}
\label{corollary:minmaxelement}
$\pi \in \AV_n(213,231)$ if and only if for $1 \leq i \leq n$,
$\pi[i:]$ start either with the maximal element
or the minimal element of $\pi[i:]$.
\end{corollary}

@@

\begin{lemma}
\label{lemma:MatchStripeToPermutation}
Let $\pi$ and $\sigma$ be two $(213,231)$-avoiding permutations,
and $t$ be a subsequence of $\pi$ of length $|\sigma|$ such that $s_t = s_\sigma$.
Then, $t$ is an occurrence of $\sigma$ in $\pi$ if and only if
$s_t$ occurs as a subsequence in $s_\pi$.
\end{lemma}

\begin{proof}
  The forward direction is obvious.
  We prove the backward direction by induction on the size of the motif
  $\sigma$.
  The base case is a motif of size $2$.
  Suppose first that $\sigma = 12$ and hence $s_\sigma = a$.
  Let $t = \pi_{i_1}\pi_{i_2}$, $i_1 < i_2$, be a subsequence of $\pi$
  such that $s_t$ occurs as a subsequence in $s_\pi$.
  Since $s_\sigma = s_t = a$, this reduces to saying that
  $\pi_{i_1} < \pi_{i_2}$, and hence $t$ is an occurrence of $\sigma = 12$ in $\pi$.
  A similar argument shows that the lemma holds true for $\sigma = 21$.
  Now, assume that the lemma is true for all motifs up to size $k \geq 2$.
  Let $\sigma \in \Av{k+1}{231,213}$ and
  let $t = \pi_{i_1}\pi_{i_2}\,\ldots\,\pi_{i_{k+1}}$,
  $i_1 < i_2 < \ldots < i_{k+1}$,
  be a subsequence of $\pi$ of length $k+1$ such that
  $s_t = s_\sigma$.
  Suppose that $s_t$ occurs in $s_\pi$ as a subsequence.
  Then, $s_t[2:] = s_\sigma[2:]$ occurs as a subsequence in $s_\pi[2:]$.
  But $\sigma[2:] \in \Av{k}{231,213}$ and hence,
  by the inductive hypothesis, it follows that
  $t[2:]$ is an occurrence of $\sigma[2:]$ in $\pi[2:]$.
  Moreover, if $s_\sigma[1] = a$ (resp. $s_\sigma[1] = d$)
  then $\sigma[1]$ is the minimum (resp. maximum) element of $\sigma$
  (Lemma~\ref{lemma:first element is 1 or n}),
  and hence, since $s_t = s_\sigma$ occurs as a subsequence in $s_\pi$,
  $t[1]$ is the minimum (resp. maximum) element of $t$.
  Therefore, $t$ is an occurrence of $\sigma$ in $\pi$.
\qed
\end{proof}

@@

\begin{proof}
According to Lemma~\ref{lemma:MatchStripeToPermutation} the problem reduces
to deciding whether $s_\sigma$ occurs as a subsequence in $s_\pi$.
A straightforward greedy approach solves this issue in linear-time.
Furthermore, in the light of Corollary~\ref{corollary:minmaxelement},
the effective construction of the two stripes is not required,
so it is an online algorithm.
\qed
% To solve the permutation pattern for a permutation in $\AV_n(213,231)$,
% we find a find a match of the stripe of the pattern in the stripe of the text.
% This can be done linearly by a greedy algorithm,
% by matching a step whenever it is possible.
% And thank to the definition of the definition of a stripe in
% Corollary~\ref{corollary:minmaxelement},
% we do not have to compute the stripe in advance.
% Therefore we have an linear (online) algorithm.
\end{proof}

@@

This section focuses on the pattern containment problem
in case only the pattern $\sigma$ avoids $231$ and $213$.
We need to consider a specific decomposition $\factora$ of $\sigma$ into factors.
For $\sigma \in S_k$, the factorization
$\sigma = \factor{\pmotif}{p} \factor{\pmotif}{p-1} \ldots \factor{\pmotif}{1}$ is defined as follows:
(i) $\sigma[k]$ is the last letter of $\factor{\pmotif}{1}$,
(ii) for every $1 \leq i < k$,
if $s_\sigma[i] = s_\sigma[i+1]$ then
$\sigma[i]\sigma[i+1]$ is part of the same factor.
For every factor $\factor{\pmotif}{j}$ of the decomposition, we let
$\first{\pmotif}{j}$ stand for the leftmost letter of $\factor{\pmotif}{j}$,
and $\firsti{\pmotif}{j}$ stand for the index of the leftmost letter of $\factor{\pmotif}{j}$.
For example,
for $\sigma = 981237654$,
we have
$s_\sigma = ddaaaddd$ and hence
$\sigma = \factor{\pmotif}{3} \factor{\pmotif}{2} \factor{\pmotif}{1}$ with
$\factor{\pmotif}{1} =7654$, $\factor{\pmotif}{2} = 123$ and $\factor{\pmotif}{3} = 98$.
Furthermore,
$\first{\pmotif}{1} = 7$, $\first{\pmotif}{2} = 1$, $\first{\pmotif}{3} = 9$,
$\firsti{\pmotif}{1} = 6$, $\firsti{\pmotif}{2} = 3$ and $\firsti{\pmotif}{3} = 1$.

@@

\begin{lemma}
\label{lemma:whereIsMax}
Given a permutation $\sigma \in \AV(213,231)$ and a suffix of its decomposition $\factor{\sigma}{i}\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$, if
$\stripe{\sigma}{\firsti{\sigma}{i}}=\ustep$ ($\stripe{\sigma}{\firsti{\sigma}{i}}=\dstep$) then the maximal (minimal) element of $\factor{\sigma}{i}\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$ is $\first{\sigma}{i-1}$.
\end{lemma}

\begin{proof}
Take the right most element of $\factor{\sigma}{i}$, this element is an ascent (descent),
so it is inferior (superior) to the next element which is $\first{\sigma}{i-1}$. Because $\factor{\sigma}{i}$ is an increasing (decreasing) subsequence, every element of $\factor{\sigma}{i}$ is inferior (superior) to $\first{\sigma}{i-1}$.
By lemma \ref{lemma:first element is 1 or n}, $\first{\sigma}{i-1}$ is the maximal (minimal) element of $\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$ thus $\first{\sigma}{i-1}$ is the maximal (minimal) element of $\factor{\sigma}{i}\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$.

\end{proof}

@@

\begin{lemma}
Given a permutation $\pmotif$ and $\factor{\pmotif}{i}$ an ascent (descent) factor.
Let's $s \in \SET{\pmotif}{\ptext}{i}{j}$, such that the element matching $\first{\pmotif}{i-1}$ is minimal (maximal), for all subsequence 
$s' \in \SET{\pmotif}{\ptext}{i}{j}$ and for all subsequence $t$ of $\ptext$, such that $t=t's'$, if $t$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$ then the subsequence $t's$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$.
\end{lemma}

\begin{proof}
Given a permutation $\pmotif$ and $\factor{\pmotif}{i}$ an ascent (descent) factor.
Let's $s \in \SET{\pmotif}{\ptext}{i}{j}$, such that the element matching $\first{\pmotif}{i-1}$ is minimal (maximal).
Let's $s' \in \SET{\pmotif}{\ptext}{i}{j}$ and $t$ subsequence of $\ptext$, such that $t=t's'$ and $t$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$ in $\pi$. 
Remark that every element of $t'$ is superior (inferior) to every element of $s'$.
Plus, remark that the element matching $\first{\pmotif}{i-1}$ in $s$ is minimal (maximal) and it is the maximal (minimal) element of the subsequence (\ref{lemma:whereIsMax}), thus
it is inferior (superior) to the element matching $\first{\pmotif}{i-1}$ in $s'$.
So every element of $t'$ is also superior (inferior) to every element of $s$. Thus the subsequence $t's$ is an occurrence of $\pmotif[\first{\pmotif}{i+1}:]$ in $\ptext$.
\end{proof}

\begin{corollary}
Let's $s \in \SET{\pmotif}{\ptext}{i}{j}$, such that $s[\first{\pmotif}{i-1}]$ is minimal (maximal). There is a match of $\pmotif$ in $\ptext$ with element $\first{\pmotif}{i}$ matched to $\ptext[j]$ if and only if there exists $t$ such that $ts$ is a match of $\pmotif$ in $\ptext$. 
\end{corollary}

@@

Consider the following problem :\\
$
\LM{\pmotif}{\ptext}{i}{j} = 
\begin{cases}
	\text{The maximal value of a subsequence} 
		& \text{If $\stripe{\sigma}{\firsti{\sigma}{i}}=\ustep$ } \\
	\text{ which is an occurence of $\sigma[\firsti{\sigma}{i}:]$ in $\pi[j:]$} & \\	
	\text{ and that minimize the maximal value} & \\
	\text{ and $\first{\sigma}{i}$ is matched with $\ptext[j]$} & \\

@@

	\text{The minimal value of a subsequence} 
		& \text{If $\stripe{\sigma}{\firsti{\sigma}{i}}=\dstep$ } \\
	\text{ which is an occurence of $\sigma[\firsti{\sigma}{i}:]$ in $\pi[j:]$} & \\	
	\text{ and that maximize the minimal value} & \\
	\text{ and $\first{\sigma}{i}$ is matched with $\ptext[j]$} & \\
\end{cases}
$

@@

\textbf{BASE :} \\
$
\LM{\pmotif}{\ptext}{1}{j} = 
\begin{cases}
		\min_{j<j'}\{\infty\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } 
			& \text{If $\stripe{\sigma}{\firsti{\sigma}{i}}=\ustep$ }  \\
		\text{ $|\factor{\pmotif}{1}| \leq LIS(j,j',\ptext[j'])  $}\}& \\
		&\\
		\max_{j<j'}\{0\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } 
			& \text{If $\stripe{\sigma}{\firsti{\sigma}{i}}=\dstep$ } \\
		\text{ $|\factor{\pmotif}{1}| \leq LDS(j,j',\ptext[j']) $}\}&\\
\end{cases}
$

\textbf{STEP :} \\
$
\LM{\pmotif}{\ptext}{i}{j} =
\begin{cases}
	\min \{\infty\} \cup  \AF{\pmotif}{\ptext}{i}{j} &
	\text{if $\stripe{\pmotif}{\first{\pmotif}{i}}=\ustep$}\\
	\max \{0\} \cup  \DF{\pmotif}{\ptext}{i}{j} &
	\text{if $\stripe{\pmotif}{\first{\pmotif}{i}}=\dstep$}\\
\end{cases}
$

@@

With $\AF{\pmotif}{\ptext}{i}{j}$ ($\DF{\pmotif}{\ptext}{i}{j}$) the set of element such that if  $\ptext[k] \in \AF{\pmotif}{\ptext}{i}{j}$ ($\ptext[k] \in \DF{\pmotif}{\ptext}{i}{j}$) then there exists a match of $\pmotif[\firsti{\pmotif}{i}:]$ in $\ptext[j:]$ and the element $\first{\pmotif}{i-1}$ is matched with element $\ptext[k]$. Given that one has the set $\SET{\pmotif}{\ptext}{i-1}{k}$, to decide if $\ptext[k] \in \AF{\pmotif}{\ptext}{i}{j}$, one needs to find an match $t$ of $\factor{\pmotif}{i}$ in $\ptext[j:k-1]$ such there exists $s \in \SET{\pmotif}{\ptext}{i-1}{k}$ and $ts$ is a match of $\pmotif[\firsti{\pmotif}{i}:]$.
Formally we define $\AF{\pmotif}{\ptext}{i}{j}$ and $\DF{\pmotif}{\ptext}{i}{j}$ such as:\\

$\AF{\pmotif}{\ptext}{i}{j}=\{j'|\text{$j<j'$ and $\LM{\pmotif}{\ptext}{i-1}{j'} \neq 0$}$ \\
\indent \indent \indent and  $|\factor{\pmotif}{i}| \leq LIS_{\ptext}(j,j'-1,\first{\pmotif}{i-1})\}$ \\

$\DF{\pmotif}{\ptext}{i}{j}=\{j'|\text{$j<j'$ and $\LM{\pmotif}{\ptext}{i-1}{j'} \neq \infty$}$ \\
\indent \indent \indent and  $|\factor{\pmotif}{i}| \leq LDS_{\ptext}(j,j'-1,\first{\pmotif}{i-1}) \}$ \\

With $LIS_{\ptext}(i,j,k)$ ($LDS_{\ptext}(i,j,k)$) is the longest increasing (decreasing) sequence in $\ptext$ starting at $i$ and ending at $j$,
with every element of this sequence
inferior (superior) to $k$.
$LIS_{\ptext}$ and $LDS_{\ptext}$ can be computed in $O(|\ptext|^2*\log(\log(|\ptext|)))$ (see \cite{Bespamyatnikh00enumeratinglongest}).\\

For the base case,
one is looking for a match for the first factor.
If the factor is an ascent (descent),
one has to find an increasing (decreasing) subsequence
in the text of same size or longest that
the size of the factor,
and to find the safest solution
one must assure that the last element
of the sequence is minimal (maximal).\\
For the induction, it follows the same idea
except that one must assure that
every element of the subsequence is superior (inferior)
to the maximum (minimal) element of the rest of the match which is given by $L$ of
the previous factor.\\

There exists a match of $\pmotif$ in $\ptext$ if and only if
there exists a $LM(n,i)\neq 0$ and $LM(n,i)\neq \infty$ for $1 \leq i \leq |\ptext|$,
with $n$ the number of factor in $\pmotif$.
Moreover the basic case can be computed in $O(n^2)$
and the induction on $O(kn^2)$ time and space.
\qed
\end{proof}

@@

We introduce property of bivincular permutation if the permutation is $(213,231)$-avoiding.

\begin{lemma}
\label{lemma:ascentDescentAscent}
If $\overline{\sigma[i]\sigma[j]}$ (thus $\sigma[j]=\sigma[i]+1$) and if $\sigma[i]$ is ascent (decent) and $\sigma[i]+1$ is ascent (decent) then $i<j$ ($j>i$) and for every $k$, $i<k<j$ ($j>k>i$), $\pi[k]$ is descent (ascent).    
\end{lemma}

\begin{proof}
Suppose that there exists $k$, $i<k<j$ ($i>k>j$),
such that $\pmotif[k]$ is ascent (descent). Ascent (descent)  elements are increasing (decreasing) so $\sigma[i]<\sigma[k]<\sigma[j]$ ($\sigma[i]<\sigma[k]<\sigma[j]$) which is in contradiction that with $\sigma[j]=\sigma[i]+1$.
\end{proof}

	\begin{proposition}
		\label{Proposition:bivincular pattern}
			Let $\sigma$ be a $(213,231)$-avoiding bivincular pattern of length $k$
			and $\pi \in S_n$.
			One can decide in $O(kn^4)$ time
			and $O(kn^4)$ space if $\pi$ contains $\sigma$.
	\end{proposition}

\begin{proof}
We consider the following problem :\\
Given a bivincular motif $\pbmotif$ with $\pmotif \in \Av{n_\pmotif}{231,213} $, and a text $\ptext \in \Av{n_\ptext}{231,213}$.
$$
\PM{\pbmotif}{\ptext}{\lb}{\ub}{i}{j}=
\begin{cases}
	true 	& \text{If $\pbmotif[i:]$ occurs in $\ptext[j:]$}\\
			& \text{with every element in $[\lb,\ub]$}\\
			& \text{and element $\pmotif[i]$ occurs in $\ptext[j]$ }\\
	false 	& otherwise\\
\end{cases}
$$

@@

The base case finds an occurrence for the right most element of the motif. If the last element do not have any restriction on position and on value, then $\PM{\pbmotif}{\ptext}{\lb}{\ub}{n_\pmotif}{j}$ is true if and only if $\pmotif[n_\pmotif]$ can occur in $\ptext[j]$. This condition is checked by assuring that $\ptext[j] \in [\lb,\ub]$.

>>>>

\begin{abstract}
	Given permutations $\sigma \in S_k$ and $\pi \in S_n$, the
  \emph{pattern containment} problem is to decide whether $\pi$ contains
  $\sigma$ as an order-isomorphic subsequence.
	We give a linear-time algorithm in case both $\pi$ and $\sigma$ avoid
	$213$ and $231$.
	For the special where only $\sigma$ avoids $213$ and $231$, we present a
	$O(kn^2)$ time algorithm. We extend our research to $(213,231)$-avoiding
	bivincular pattern and present a $O(kn^5)$ time algorithm.
\end{abstract}

@@

This section is devoted to presenting a fast algorithm for deciding if
$\sigma \preceq \pi$
in case both $\pi$ and $\sigma$ are $(213,231)$-avoiding.
We begin with an easy but crucial structure lemma.

@@

\begin{proof}
Any other initial element would serve as a `$2$' in either a
$231$ or $213$ with $1$ and $n$ as the `$1$' and `$3$' respectively.
$\qed$
\end{proof}

Notice that Lemma~\ref{lemma:first element is 1 or n} gives as an
immediate consequence that the number of $(213,231)$-avoiding $n$-permutations
is $2^{n-1}$ for $n\geq 1$ \cite{Simion:Schmidt:EJC:1985}.
The above lemma also gains in interest in the form of the following corollaries.

\begin{corollary}
\label{corollary:minmaxelement}
$\pi \in \AV_n(213,231)$ if and only if for $1 \leq i \leq n$,
$\pi[i]$ is the minimal or the maximal element of $\pi[i:]$.
\end{corollary}

\begin{corollary}
\label{corollary:max_is_ascent}
For $\pi \in \AV_n(213,231)$ if $\pi[i]$ is an ascent element then $\pi[i]$ is the minimal of $\pi[i:]$
and if if $\pi[i]$ is a descent element then $\pi[i]$ is the maximal of $\pi[i:]$
\end{corollary}

@@

\begin{lemma}
\label{lemma:MatchStripeToPermutation}
If $\pi$ and $\sigma$ be two $(213,231)$-avoiding permutations,
and $t$ be a subsequence of $\pi$ of length $|\sigma|$ such that $s_t = s_\sigma$.
Then, $t$ is an occurrence of $\sigma$ in $\pi$ if and only if
$s_t$ occurs as a subsequence in $s_\pi$.
\end{lemma}

\begin{proof}
  The forward direction is obvious.
  We prove the backward direction by induction on the size of the motif
  $\sigma$.
  The base case is a motif of size $2$.
  Suppose first that $\sigma = 12$ and hence $s_\sigma = a$.
  Let $t = \pi_{i_1}\pi_{i_2}$, $i_1 < i_2$, be a subsequence of $\pi$
  such that $s_t$ occurs as a subsequence in $s_\pi$.
  Since $s_\sigma = s_t = a$, this reduces to saying that
  $\pi_{i_1} < \pi_{i_2}$, and hence $t$ is an occurrence of $\sigma = 12$ in $\pi$.
  A similar argument shows that the lemma holds true for $\sigma = 21$.
  Now, assume that the lemma is true for all motifs up to size $k \geq 2$.
  Let $\sigma \in \Av{k+1}{231,213}$ and
  let $t = \pi_{i_1}\pi_{i_2}\,\ldots\,\pi_{i_{k+1}}$,
  $i_1 < i_2 < \ldots < i_{k+1}$,
  be a subsequence of $\pi$ of length $k+1$ such that
  $s_t = s_\sigma$.
  Suppose that $s_t$ occurs in $s_\pi$ as a subsequence.
  Then, $s_t[2:] = s_\sigma[2:]$ occurs as a subsequence in $s_\pi[2:]$.
  But $\sigma[2:] \in \Av{k}{231,213}$ and hence,
  by the inductive hypothesis, it follows that
  $t[2:]$ is an occurrence of $\sigma[2:]$ in $\pi[2:]$.
  Moreover, if $s_\sigma[1] = a$ (resp. $s_\sigma[1] = d$)
  then $\sigma[1]$ is the minimum (resp. maximum) element of $\sigma$
  (Lemma~\ref{lemma:first element is 1 or n}),
  and hence, since $s_t = s_\sigma$ occurs as a subsequence in $s_\pi$,
  $t[1]$ is the minimum (resp. maximum) element of $t$.
  Therefore, $t$ is an occurrence of $\sigma$ in $\pi$.
$\qed$
\end{proof}

@@

\begin{proof}
According to Lemma~\ref{lemma:MatchStripeToPermutation} the problem reduces
to deciding whether $s_\sigma$ occurs as a subsequence in $s_\pi$.
A straightforward greedy approach solves this issue in linear-time.
Furthermore, in the light of Corollary~\ref{corollary:minmaxelement},
the effective construction of the two stripes is not required,
so it is an online algorithm.
$\qed$
% To solve the permutation pattern for a permutation in $\AV_n(213,231)$,
% we find a find a match of the stripe of the pattern in the stripe of the text.
% This can be done linearly by a greedy algorithm,
% by matching a step whenever it is possible.
% And thank to the definition of the definition of a stripe in
% Corollary~\ref{corollary:minmaxelement},
% we do not have to compute the stripe in advance.
% Therefore we have an linear (online) algorithm.
\end{proof}

@@

This section focuses on the pattern containment problem
in case only the pattern $\sigma$ avoids $231$ and $213$.
We need to consider a specific decomposition $\factora$ of $\sigma$ into factors.
For $\sigma \in S_k$, the factorization
$\sigma = \factor{\pmotif}{p} \factor{\pmotif}{p-1} \ldots \factor{\pmotif}{1}$ is defined as follows:
(i) $\sigma[k]$ is the last letter of $\factor{\pmotif}{1}$,
(ii)for every $1 \leq i < k$,
if $s_\sigma[i] = s_\sigma[i+1]$ then
$\sigma[i]\sigma[i+1]$ is part of the same factor.
For every factor $\factor{\pmotif}{j}$ of the decomposition, we let
$\first{\pmotif}{j}$ stand for the leftmost letter of $\factor{\pmotif}{j}$,
and $\firsti{\pmotif}{j}$ stand for the index of the leftmost letter of $\factor{\pmotif}{j}$.
For example,
for $\sigma = 981237654$,
we have
$s_\sigma = ddaaaddd$ and hence
$\sigma = \factor{\pmotif}{3} \factor{\pmotif}{2} \factor{\pmotif}{1}$ with
$\factor{\pmotif}{1} =7654$, $\factor{\pmotif}{2} = 123$ and $\factor{\pmotif}{3} = 98$.
Furthermore,
$\first{\pmotif}{1} = 7$, $\first{\pmotif}{2} = 1$, $\first{\pmotif}{3} = 9$,
$\firsti{\pmotif}{1} = 6$, $\firsti{\pmotif}{2} = 3$ and $\firsti{\pmotif}{3} = 1$.

@@

\begin{lemma}
\label{lemma:whereIsMax}
Given a permutation $\sigma \in \AV(213,231)$ and a suffix of its decomposition $\factor{\sigma}{i}\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$, if
$\firsti{\sigma}{i}$ is an ascent element (respectively  if $\firsti{\sigma}{i}$ is a descent element) then the maximal (resp. minimal) element of $\factor{\sigma}{i}\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$ is $\first{\sigma}{i-1}$.
\end{lemma}

\begin{proof}
Take the right most element of $\factor{\sigma}{i}$, if this element is an ascent element,
then it is smaller than the next element which is $\first{\sigma}{i-1}$. Because $\factor{\sigma}{i}$ is an increasing subsequence, every element of $\factor{\sigma}{i}$ is inferior to $\first{\sigma}{i-1}$.
By lemma \ref{lemma:first element is 1 or n}, $\first{\sigma}{i-1}$ is the maximal element of $\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$ thus $\first{\sigma}{i-1}$ is the maximal element of $\factor{\sigma}{i}\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$.
We use a similar argument if the rightmost element is a decent element.
$\qed$
\end{proof}

@@

\begin{lemma}
Given a permutation $\pmotif$ and $\factor{\pmotif}{i}$ an ascent (respectively descent) factor.
If $s \in \SET{\pmotif}{\ptext}{i}{j}$, such that the element matching $\first{\pmotif}{i-1}$ is minimal (resp. maximal), for all subsequence 
$s' \in \SET{\pmotif}{\ptext}{i}{j}$ and for all subsequence $t$ of $\ptext$, such that $t=t's'$, if $t$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$ then the subsequence $t's$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$.
\end{lemma}

\begin{proof}
By definition $s$ is an occurrence of $\pmotif[\firsti{\pmotif}{i}:]$. To prove that $ts$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$ we prove that every element of t is larger than every element of $s$. If $ts'$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$ so every element of $t$ is larger than every element of $s'$. Moreover the maximal element of $s$ is smaller than  the maximal element of $s'$ so every element of $s$ is smaller than every element of $s'$ thus every element of $s$ is smaller than every element of $t$. We use a similar argument if $\factor{\pmotif}{i}$ a descent factor.
$\qed$
\end{proof}

\begin{corollary}
Given a permutation $\pmotif$ and $\factor{\pmotif}{i}$ an ascent (respectively descent) factor.
If $s \in \SET{\pmotif}{\ptext}{i}{j}$, such that $s[\first{\pmotif}{i-1}]$ is minimal (resp. maximal). There is a match of $\pmotif$ in $\ptext$ with element $\first{\pmotif}{i}$ matched to $\ptext[j]$ if and only if there exists $t$ such that $ts$ is a match of $\pmotif$ in $\ptext$. 
\end{corollary}

@@

Consider the following problem :\\
$
\LM{\pmotif}{\ptext}{i}{j} = 
\begin{cases}
	\text{The maximal value of a subsequence} 
		& \text{If $\first{\sigma}{i}$ is  } \\
	\text{ which is an occurence of $\sigma[\firsti{\sigma}{i}:]$ in $\pi[j:]$} & 
		\text{ an ascent element } \\	
	\text{ and that minimize the maximal value} & \\
	\text{ and $\first{\sigma}{i}$ is matched with $\ptext[j]$} & \\

@@

	\text{The minimal value of a subsequence} 
		& \text{If $\first{\sigma}{i}$ is  } \\
	\text{ which is an occurence of $\sigma[\firsti{\sigma}{i}:]$ in $\pi[j:]$} & \text{ a descent element } \\	
	\text{ and that maximize the minimal value} & \\
	\text{ and $\first{\sigma}{i}$ is matched with $\ptext[j]$} & \\
\end{cases}
$

@@

\textbf{BASE :} \\
$
\LM{\pmotif}{\ptext}{1}{j} = 
\begin{cases}
		\min_{j<j'}\{\infty\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } 
			& \text{If $\first{\sigma}{i}$ is}  \\
		\text{ $|\factor{\pmotif}{1}| \leq LIS(j,j',\ptext[j'])  $}\}&\text{ an ascent element } \\
		&\\
		\max_{j<j'}\{0\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } 
			& \text{If $\first{\sigma}{i}$ is}  \\
		\text{ $|\factor{\pmotif}{1}| \leq LDS(j,j',\ptext[j']) $}\}&\text{ a descent element }\\
\end{cases}
$

\textbf{STEP :} \\
$
\LM{\pmotif}{\ptext}{i}{j} =
\begin{cases}
	\min \{\infty\} \cup  \AF{\pmotif}{\ptext}{i}{j} &
	\text{If $\first{\sigma}{i}$ is an ascent element}\\
	\max \{0\} \cup  \DF{\pmotif}{\ptext}{i}{j} &
	\text{If $\first{\sigma}{i}$ is  a descent element}\\
\end{cases}
$

@@

With $\AF{\pmotif}{\ptext}{i}{j}$ and $\DF{\pmotif}{\ptext}{i}{j}$ the set of element such that if  $\ptext[k] \in \AF{\pmotif}{\ptext}{i}{j}$ or if $\ptext[k] \in \DF{\pmotif}{\ptext}{i}{j}$ then there exists a match of $\pmotif[\firsti{\pmotif}{i}:]$ in $\ptext[j:]$ and the element $\first{\pmotif}{i-1}$ is matched with element $\ptext[k]$. Given that one has the set $\SET{\pmotif}{\ptext}{i-1}{k}$, to decide if $\ptext[k] \in \AF{\pmotif}{\ptext}{i}{j}$, one needs to find an occurrence $t$ of $\factor{\pmotif}{i}$ in $\ptext[j:k-1]$ such there exists $s \in \SET{\pmotif}{\ptext}{i-1}{k}$ and $ts$ is a match of $\pmotif[\firsti{\pmotif}{i}:]$. 
Formally we define $\AF{\pmotif}{\ptext}{i}{j}$ and $\DF{\pmotif}{\ptext}{i}{j}$ such as:\\

$\AF{\pmotif}{\ptext}{i}{j}=\{\ptext[j']|\text{$j<j'$ and $\LM{\pmotif}{\ptext}{i-1}{j'} \neq 0$}$ \\
\indent \indent \indent and  $|\factor{\pmotif}{i}| \leq LIS_{\ptext}(j,j'-1,\first{\pmotif}{i-1})\}$ \\

$\DF{\pmotif}{\ptext}{i}{j}=\{\ptext[j']|\text{$j<j'$ and $\LM{\pmotif}{\ptext}{i-1}{j'} \neq \infty$}$ \\
\indent \indent \indent and  $|\factor{\pmotif}{i}| \leq LDS_{\ptext}(j,j'-1,\first{\pmotif}{i-1}) \}$ \\

With $LIS_{\ptext}(i,j,k)$ (respectively $LDS_{\ptext}(i,j,k)$) is the longest increasing (resp. decreasing) sequence in $\ptext$ starting at $i$ and ending at $j$,
with every element of this sequence
smaller (resp. bigger) than $k$.
$LIS_{\ptext}$ and $LDS_{\ptext}$ can be computed in $O(|\ptext|^2\log(\log(|\ptext|)))$ (see \cite{Bespamyatnikh00enumeratinglongest}).\\

For the base case,
one is looking for a match for the first factor.
If the factor is an ascent (respectively descent) factor, then
one has to find an increasing (resp. decreasing) subsequence
in the text of same size or longest that
the size of the first factor.
To find the "safest" solution
one must assure that the right most element
of the sequence is minimal (resp. maximal).\\
For the induction, it follows the same idea
except that one must assure that
every element of the subsequence is larger (resp. smaller)
than the maximum (resp. minimal) element of the rest of the match which is given by the inductive problem of
the previous factor.\\

There exists a match of $\pmotif$ in $\ptext$ if and only if
there exists a $LM(n,i)\neq 0$ and $LM(n,i)\neq \infty$ for $1 \leq i \leq |\ptext|$,
with $n$ the number of factor in $\pmotif$.
Moreover the basic case can be computed in $O(n^2)$
and the induction on $O(kn^2)$ time and space.
$\qed$
\end{proof}

@@

We introduce property of bivincular permutation for a motif avoiding both $213$ and $231$.

\begin{lemma}
\label{lemma:ascentDescentAscent}
If $\overline{\sigma[i]\sigma[j]}$ (thus $\sigma[j]=\sigma[i]+1$) and if $\sigma[i]$ is an ascent (respectively decent) element and $\sigma[i]+1$ is an ascent (resp. decent) element then $i<j$ ($j>i$) and for every $k$, $i<k<j$ ($j>k>i$), $\pi[k]$ is a descent (resp. ascent) element.    
\end{lemma}

\begin{proof}
Suppose that there exists $k$, $i<k<j$,
such that $\pmotif[k]$ is ascent. Ascent elements are increasing  so $\sigma[i]<\sigma[k]<\sigma[j]$ which is in contradiction that with $\sigma[j]=\sigma[i]+1$.
We use a similar argument if $\sigma[i]$ is a descent element
$\qed$
\end{proof}

	\begin{proposition}
		\label{Proposition:bivincular pattern}
			Let $\sigma$ be a bivincular motif avoiding $213$ and $231$ of length $k$
			and $\pi \in S_n$.
			One can decide in $O(kn^5)$ time
			and $O(kn^5)$ space if $\sigma$ occurs in $\pi$.
	\end{proposition}

\begin{proof}
We consider the following problem :\\
Given a bivincular motif $\pbmotif$ with $\pmotif \in \Av{n_\pmotif}{231,213} $, and a text $\ptext \in \Av{n_\ptext}{231,213}$.
$$
\PM{\pbmotif}{\ptext}{\lb}{\ub}{i}{j}=
\begin{cases}
	true 	& \text{If $\pbmotif[i:]$ occurs in $\ptext[j:]$}\\
			& \text{with every element is in $[\lb,\ub]$}\\
			& \text{and element $\pmotif[i]$ occurs in $\ptext[j]$ }\\
	false 	& otherwise\\
\end{cases}
$$

@@

The base case finds an occurrence for the right most element of the motif. If the last element does not have any restriction on position and on value, then $\PM{\pbmotif}{\ptext}{\lb}{\ub}{n_\pmotif}{j}$ is true if and only if $\pmotif[n_\pmotif]$ can occur in $\ptext[j]$. This condition is checked by assuring that $\ptext[j] \in [\lb,\ub]$.

\textbf{STEP :} \\
For clarity of reading we consider 3 cases for the problem $\PM{\pbmotif}{\ptext}{\lb}{\ub}{i}{j}$ :
\begin{itemize}
	\item If $\ptext[j] \notin [\lb,\ub]$ then :
	$$
	\PM{\pbmotif}{\ptext}{\lb}{\ub}{i}{j} = false
	$$
	this is straightforward with the definition. If $\ptext[j] \notin [\lb,\ub]$ then it can not be part of an occurrence of  $\pmotif[i:]$ in $\ptext[i:]$ with every element is in $[\lb,\ub]$.
